---
title: "Mini Data Analysis Milestone 2"
output: github_document
---

*To complete this milestone, you can either edit [this `.rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are commented out with `<!--- start your work here--->`. When you are done, make sure to knit to an `.md` file by changing the output in the YAML header to `github_document`, before submitting a tagged release on canvas.*

# Welcome to the rest of your mini data analysis project!

In Milestone 1, you explored your data. and came up with research questions. This time, we will finish up our mini data analysis and obtain results for your data by: 

- Making summary tables and graphs 
- Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

We will also explore more in depth the concept of *tidy data.* 

**NOTE**: The main purpose of the mini data analysis is to integrate what you learn in class in an analysis. Although each milestone provides a framework for you to conduct your analysis, it's possible that you might find the instructions too rigid for your data set. If this is the case, you may deviate from the instructions -- just make sure you're demonstrating a wide range of tools and techniques taught in this class.

# Instructions

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work here--->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to your mini-analysis GitHub repository, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 50 points: 45 for your analysis, and 5 for overall reproducibility, cleanliness, and coherence of the Github submission. 

**Research Questions**: In Milestone 1, you chose two research questions to focus on. Wherever realistic, your work in this milestone should relate to these research questions whenever we ask for justification behind your work. In the case that some tasks in this milestone don't align well with one of your research questions, feel free to discuss your results in the context of a different research question.

# Learning Objectives

By the end of this milestone, you should:

-   Understand what *tidy* data is, and how to create it using `tidyr`.
-   Generate a reproducible and clear report using R Markdown.
-   Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

# Setup

Begin by loading your data and the tidyverse package below:

```{r, message = FALSE}
library(datateachr) # <- might contain the data you picked!
library(tidyverse)
```

# Task 1: Process and summarize your data 

From milestone 1, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions. 

### 1.1 (1 point) 

First, write out the 4 research questions you defined in milestone 1 were. This will guide your work through milestone 2:

<!-------------------------- Start your work below ---------------------------->
1. Is there a flow trend for each year (ie. how does the flow change throughout a year)?
2. Is there a trend for the amount of missing values in the variable SYM?
3. Are there years or months where flow is more volatile?
4. How do the maximum and minimum flow rates (sub-setting by the extreme variable) change throughout the years/months?
<!----------------------------------------------------------------------------->

Here, we will investigate your data using various data manipulation and graphing functions.

### 1.2 (8 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question.

Ensure that the output of each operation is printed!

Also make sure that you're using dplyr and ggplot2 rather than base R. Outside of this project, you may find that you prefer using base R functions for certain tasks, and that's just fine! But part of this project is for you to practice the tools we learned in class, which is dplyr and ggplot2.

**Summarizing:**

1.  Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2.  Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3.  Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4. Compute the proportion and counts in each category of one categorical variable across the groups of another categorical variable from your data. Do not use the function `table()`!

**Graphing:**

6. Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.
7. Make a graph where it makes sense to customize the alpha transparency.

Using variables and/or tables you made in one of the "Summarizing" tasks: 

8. Create a graph that has at least two geom layers. 
9. Create 3 histograms, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->
```{r}
data <- flow_sample
df1 <- as.data.frame(data)

# For Question 1: I Chose to do task 2 and 8

# Determine the observation counts of the month variable using group_by to group unique values and summarize to calculate the observations of each values. This will tell us whether some months may be more or less accurate due to sample size differences
result1 <- data %>%
  group_by(month) %>%
  summarise(Count = n())
print(result1)

# Make two geom layers (scatter plot and linear regression line). This would give us a rough idea of what the flow trend is like over the years. Here I set the scatter plot size to be blue and a size of 3 whereas for the linear regression line, I use linear modeling with a red dashed line.
plot1 <- ggplot(data, aes(x = year, y = flow)) +
  geom_point(color = "blue", size = 3) +
  geom_smooth(method = "lm", color = "red", linetype = "dashed")
print(plot1)


# For Question 2: I Chose to do task 2 and 8
# Group the data by the year column and calculate the number of NA values in the sym column. Then summarize the sum of the NA values.
result2 <- data %>%
  group_by(year) %>%
  summarise(NA_count = sum(is.na(sym)))
print(result2)
# Plot the results of result2 (ie. the amount of NA values in sym column for each year). I chose to use geom_point and geom_line as they allow better visualization of where there is a decrease or increase in the amount of sym NA values since there are so many years. Here I chose a solid blue line for the line graph and for the scatter plot, I chose red dots with a size of 4 and a filled circle shape.
plot2 <- ggplot(result2, aes(x = year)) +
  geom_line(aes(y = NA_count), color = "blue", linetype = "solid", size = 1.5) +
  geom_point(aes(y = NA_count), color = "red", shape = 16, size = 4) 

print(plot2)

# For Question 3: I chose to do task 1 and 8
# Here I calculate the summary stats for the flow variable across the years. First, I group by years (using group_by) and then I  calculate the different summary statistics (ie. range, mean, median, SD, and IQR) all using the flow variable since that's what we're interested in.
result3 <- data %>%
  group_by(year) %>%
  summarise(
    Range = max(flow, na.rm = TRUE) - min(flow, na.rm = TRUE),
    Mean = mean(flow, na.rm = TRUE),
    Median = median(flow, na.rm = TRUE),
    SD = sd(flow, na.rm = TRUE),
    IQR = IQR(flow, na.rm = TRUE)
  )
print(result3)

# For plotting the volatility of the flow over the years, I chose to use a bar plot and error bars to look at the range and SD respectively. Specifically for the error bars, I chose to show the standard deviations using the formula mean +/- SD in red.

plot3 <- ggplot(result3, aes(x = year)) +
  geom_bar(aes(y = Range), stat = "identity", fill = "skyblue", alpha = 0.7) +  # First geom layer: Bar plot for Range
  geom_errorbar(aes(ymin = Mean - SD, ymax = Mean + SD), width = 0.2, color = "red", size = 1.5) +  # Second geom layer: Error bars for Standard Deviation
  labs(y = "Flow", title = "Volatility of Flow Across Years")
print(plot3)


# For Question 4: I chose to do task 2 and 8
# Create a new df where the max_flow column contains the max flow rate for each year and the min flow rate is the minimum flow rate of each year. Once again, I grouped by year.
result4 <- data %>%
  group_by(year) %>%
  summarise(
    max_flow = max(flow, na.rm = TRUE),
    min_flow = min(flow, na.rm = TRUE)
  )
print(result4)

# Here I created two line plots visualizing the max and min flow rate from result4. 
plot4 <- ggplot(result4, aes(x = year)) +
  geom_line(aes(y = max_flow), color = "blue", linetype = "solid", size = 1) +
  geom_line(aes(y = min_flow), color = "red", linetype = "solid", size = 1) +
  labs(y = "Flow Rate", title = "Maximum and Minimum Flow Rates Over the Years") +
  theme_minimal()
print(plot4)
```

<!----------------------------------------------------------------------------->

### 1.3 (2 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!------------------------- Write your answer here ---------------------------->
Given the operations completed above, we now have a better understanding of the overall trends of the sample data. By examining the flow trends thorughout the year, patterns and seasonal changes in flow rates are identified. Assessing trends in the missing values over time also highlights potential data collection issues and periods of incomplete information. Moreover, analyzing flow volatitiy across diffrent years and months offers crucial information for downstream analyses (eg. questions 1,3, and 4). I think my research questions doesn't necessarily need to be refined.

<!----------------------------------------------------------------------------->

# Task 2: Tidy your data 

In this task, we will do several exercises to reshape our data. The goal here is to understand how to do this reshaping with the `tidyr` package.

A reminder of the definition of *tidy* data:

-   Each row is an **observation**
-   Each column is a **variable**
-   Each cell is a **value**

### 2.1 (2 points)

Based on the definition above, can you identify if your data is tidy or untidy? Go through all your columns, or if you have \>8 variables, just pick 8, and explain whether the data is untidy or tidy.

<!--------------------------- Start your work below --------------------------->
I would argue that my data is tidy since the data fits the format of "tidy" data (ie. each row is an observation, each column is a variable and each cell is a value). In particular, each row represents the flow data from a station at a given timepoint. I would argue that even the NA cells in the sym column because NA are technically values which represent missing data.

<!----------------------------------------------------------------------------->

### 2.2 (4 points)

Now, if your data is tidy, untidy it! Then, tidy it back to it's original state.

If your data is untidy, then tidy it! Then, untidy it back to it's original state.

Be sure to explain your reasoning for this task. Show us the "before" and "after".

<!--------------------------- Start your work below --------------------------->
```{r}
# Before (Making it untidy)
untidy_data <- as.data.frame(t(data))
untidy_data

# After (Making it back to tidy)
tidy_data <- as.data.frame(t(untidy_data),row.names = FALSE)
tidy_data

# Given the definition of tidy data, by transposing our data, our data would inherently be untidy. Here, each row is a variable and each column is an observation.
```

<!----------------------------------------------------------------------------->

### 2.3 (4 points)

Now, you should be more familiar with your data, and also have made progress in answering your research questions. Based on your interest, and your analyses, pick 2 of the 4 research questions to continue your analysis in the remaining tasks:

<!-------------------------- Start your work below ---------------------------->
Here are the 2 research questions that I want to pursue
1. Is there a flow trend for each year (ie. how does the flow change throughout a year)?
2. Are there years or months where flow is more volatile?

<!----------------------------------------------------------------------------->

Explain your decision for choosing the above two research questions.

<!--------------------------- Start your work below --------------------------->
I still think the flow trends are a very interesting point to address and would like to continue exploring the temporal impact on it, but also the temporal impact on the flow's volatility.
<!----------------------------------------------------------------------------->

Now, try to choose a version of your data that you think will be appropriate to answer these 2 questions. Use between 4 and 8 functions that we've covered so far (i.e. by filtering, cleaning, tidy'ing, dropping irrelevant columns, etc.).

(If it makes more sense, then you can make/pick two versions of your data, one for each research question.) 

<!--------------------------- Start your work below --------------------------->

```{r}
# Here I filter out the rows where flow has NA values and then remove the station_id and sym columns (using the select function) as they are not useful for my analysis. I then use rename to make the column names better (ie. capitalize the first letter) and finally I grouped by year.
updated_data <- data %>%
  filter(!is.na(flow)) %>%  # Remove rows with NA values in the flow column
  select(-station_id, -sym) %>%  # Remove station_id and sym columns
  rename(Year = year, Month = month, Day = day, FlowRate = flow) %>%  # Rename columns
  group_by(Year)

updated_data
```
            
<!----------------------------------------------------------------------------->

# Task 3: Modelling

## 3.0 (no points)

Pick a research question from 1.2, and pick a variable of interest (we'll call it "Y") that's relevant to the research question. Indicate these.

<!-------------------------- Start your work below ---------------------------->

**Research Question**: Is there a flow trend for each year (ie. how does the flow change throughout a year)?

**Variable of interest**: Flow Rate 

<!----------------------------------------------------------------------------->

## 3.1 (3 points)

Fit a model or run a hypothesis test that provides insight on this variable with respect to the research question. Store the model object as a variable, and print its output to screen. We'll omit having to justify your choice, because we don't expect you to know about model specifics in STAT 545.

-   **Note**: It's OK if you don't know how these models/tests work. Here are some examples of things you can do here, but the sky's the limit.

    -   You could fit a model that makes predictions on Y using another variable, by using the `lm()` function.
    -   You could test whether the mean of Y equals 0 using `t.test()`, or maybe the mean across two groups are different using `t.test()`, or maybe the mean across multiple groups are different using `anova()` (you may have to pivot your data for the latter two).
    -   You could use `lm()` to test for significance of regression coefficients.

<!-------------------------- Start your work below ---------------------------->
```{r}
# Use the linear regression model function lm() to fit linear regression model to look at the relationship between flow rate and year
flow_model <- lm(FlowRate ~ Year, data = updated_data)
summary(flow_model)
```

<!----------------------------------------------------------------------------->

## 3.2 (3 points)

Produce something relevant from your fitted model: either predictions on Y, or a single value like a regression coefficient or a p-value.

-   Be sure to indicate in writing what you chose to produce.
-   Your code should either output a tibble (in which case you should indicate the column that contains the thing you're looking for), or the thing you're looking for itself.
-   Obtain your results using the `broom` package if possible. If your model is not compatible with the broom function you're needing, then you can obtain your results by some other means, but first indicate which broom function is not compatible.

<!-------------------------- Start your work below ---------------------------->
```{r}
# Here i am making a tibble which would have the info from the regression model (eg. p values, std error etc.). I am using this to assess the relationship between time (year) and the flow rate.

library(broom)
# Extract regression coefficients and statistics (with tidy() from broom) then look at the model summary
flow_model_summary <- tidy(flow_model)
flow_model_summary
```

<!----------------------------------------------------------------------------->

# Task 4: Reading and writing data

Get set up for this exercise by making a folder called `output` in the top level of your project folder / repository. You'll be saving things there.

## 4.1 (3 points)

Take a summary table that you made from Task 1, and write it as a csv file in your `output` folder. Use the `here::here()` function.

-   **Robustness criteria**: You should be able to move your Mini Project repository / project folder to some other location on your computer, or move this very Rmd file to another location within your project repository / folder, and your code should still work.
-   **Reproducibility criteria**: You should be able to delete the csv file, and remake it simply by knitting this Rmd file.

<!-------------------------- Start your work below ---------------------------->
```{r}
write.csv(result1, file = here::here("output", "result1.csv"), row.names = FALSE)
```

<!----------------------------------------------------------------------------->

## 4.2 (3 points)

Write your model object from Task 3 to an R binary file (an RDS), and load it again. Be sure to save the binary file in your `output` folder. Use the functions `saveRDS()` and `readRDS()`.

-   The same robustness and reproducibility criteria as in 4.1 apply here.

<!-------------------------- Start your work below ---------------------------->
```{r}
saveRDS(flow_model, file = here::here("output", "flow_model.rds"))
```

<!----------------------------------------------------------------------------->

# Overall Reproducibility/Cleanliness/Coherence Checklist 

Here are the criteria we're looking for.

## Coherence (0.5 points)

The document should read sensibly from top to bottom, with no major continuity errors. 

The README file should still satisfy the criteria from the last milestone, i.e. it has been updated to match the changes to the repository made in this milestone. 

## File and folder structure (1 points)

You should have at least three folders in the top level of your repository: one for each milestone, and one output folder. If there are any other folders, these are explained in the main README.

Each milestone document is contained in its respective folder, and nowhere else.

Every level-1 folder (that is, the ones stored in the top level, like "Milestone1" and "output") has a `README` file, explaining in a sentence or two what is in the folder, in plain language (it's enough to say something like "This folder contains the source for Milestone 1").

## Output (1 point)

All output is recent and relevant:

-   All Rmd files have been `knit`ted to their output md files. 
-   All knitted md files are viewable without errors on Github. Examples of errors: Missing plots, "Sorry about that, but we can't show files that are this big right now" messages, error messages from broken R code
-   All of these output files are up-to-date -- that is, they haven't fallen behind after the source (Rmd) files have been updated.
-   There should be no relic output files. For example, if you were knitting an Rmd to html, but then changed the output to be only a markdown file, then the html file is a relic and should be deleted.

Our recommendation: delete all output files, and re-knit each milestone's Rmd file, so that everything is up to date and relevant.

## Tagged release (0.5 point)

You've tagged a release for Milestone 2. 

### Attribution

Thanks to Victor Yuan for mostly putting this together.
